The Heavy Photon Search Test Detector

### 2013-10-26 hps-testrun-nim.tex
Fixing spaces after macros.

This paper reviews the HPS Test Run apparatus. It documents the successful performance of the trigger, data 
acquisition, silicon tracker, and ECal and
shows that the performance assumed in calculating the physics reach of the experiment is realistic.
Of particular importance, data from the dedicated photon beam running has been used to compare
the measured trigger rates with those expected in simulation. The Test Run trigger rate is almost entirely due to
photons which have converted to \ee pairs in a thin target upstream of HPS and it is sensitive to the particulars 
of their multiple Coulomb scattering in the conversion target.  Since, during electron beam running, the tails of 
the multiply Coulomb scattered primary beam are the dominant source of occupancy in the tracker and the 
trigger rate in the Ecal, they must be understood quantitatively. The trigger rates in the Test Run are also 
sensitive to the tails of the multiple Coulomb scattering distribution. Good agreement between the measured 
and simulated rates confirms our understanding of multiple Coulomb scattering and 
the reliability of the background simulation used to benchmark the physics reach of the HPS experiment.

@@

The HPS Test apparatus was designed as a 
first step of a multi-stage approach 
to the HPS experiment. The primary goal was to demonstrate that the technology and design approach 
chosen are sound for running in such a challenging environment. Nevertheless, a secondary but 
important aspect was that, after proving the technical concepts,  the apparatus would have the 
capability to discover \Aprime events. The implication was that the design of the HPS Test apparatus 
would be subject to almost the same constraints and requirements as the full experiment.
The kinematics of \Aprime production typically results in final state particles within a few degrees of the incoming beam, especially at low 
$m_{\Aprime}$.  
%Step
This implies that the apparatus for an \Aprime search in electron interactions with fix targets must operate in 
close proximity to an intense electron beam that passes through a high-Z target. 
Hence it should cope with high backgrounds and tolerate the high radiation environment. 
To meet this challenge, HPS will deploy high rate and radiation hard detector composed of 
a silicon vertex tracker with 40~MHz readout and a PbWO$_4$ crystal calorimeter with 250~MHz 
readout, capable of triggering multi-prong events at top to 50~kHz rate.

@@

The principle purpose of the HPS Test Silicon Vertex Tracker (SVT) is the efficient detection of 
charged particles and reconstruction of their momentum and angles. The precise measurements 
forms the primary input to the extraction of a signal from \Aprime decays; both from the precise 
measurement of the invariant mass of its \ee decay products and the position of its decay vertex 
downstream of the target. Track reconstruction is also important for energy scale calibration of the 
electromagnetic calorimeter due to the better momentum resolution of tracks compared to clusters 
in the calorimeter.

\subsection{SVT Requirements and Constraints}
%The primary goal was to demonstrate that the technology and design approach 
%chosen for the SVT was sound for running in the challenging environment at JLab. 
%Nevertheless, a secondary but 
%important aspect was that, after proving the technical concepts,  the apparatus would have the 
%capability to discover \Aprime events. The implication was that the design of the SVT in the 
%HPS Test apparatus would be subject to almost the same constraints and requirements as the full 
%experiment except for certain aspects regarding longevity and robustness subject to cost, beam time 
%and construction and design time constraints for the Test run.

@@

It is this last challenge that has the most interesting and obvious impacts on the design of the SVT. To 
achieve good vertexing performance, the first layer must be placed no more than about 10 cm downstream 
of the target. At that distance, it is found that the active region of a sensor can be placed as close to 1.5 mm 
from the center of the beam, defining a 15 mrad ``dead zone" in the detector around the plane of the 
primary beam. At the edge of this ``dead zone," the radiation dose approaches $10^{15}$ electrons/cm
$^2$/month, or roughly $3 \times 10^{14}$~\fluenceunit/month~\cite{Rashevskaya:2002nd}, as shown in 
Fig.~\ref{fig:radiation} and~\ref{fig:rates}.
%===================
\begin{figure}[]
\includegraphics[width=7cm]{figures/radiation}
\label{fig:radiation}
\caption{\small Simulation of the fluence 10~cm downstream of the target in the active region of the SVT ($|Y|>0.15~$cm) looking downstream; electrons bend towards $+x$.}
\end{figure}
%===================
%===================
\begin{figure}[]
\includegraphics[width=7cm]{figures/rates}
\label{fig:rates}
\caption{\small Simulation of the occupancy in the SVT 10~cm downstream of the target as a function of the 
vertical distance from the beam center.}
\end{figure}
%=================== 
This requires the sensors to be actively cooled. Meanwhile, very low-energy delta rays from beam gas interactions multiply the density of background hits, so the SVT must operate inside the beam vacuum.  Finally, in order to protect the sensors, the detector must be movable so that it can be retracted during periods of uncertain beam conditions.  
%To summarize the SVT must:
%\begin{itemize}
%\item have a uniform acceptance in the forward region; up to $\theta_y=XXXXmrad$ and as close as 
%possible to the beam axis to catch boosted decay products,
%\item >95\% reconstruction efficiency for electron and positrons  with energy down to $XXXX$~MeV,
%\item momentum resolution of XXXX\% to separate a promptly decaying \Aprime signal from the irreducible background,
%\item vertex resolution $<XXXX$~mm along the beam line direction to search for displaced \Aprime decays,
%\item event timing resolution down to 2~ns, 
%\item sustain a occupancy of up to 1\%, trigger rates of $\approx50$~kHz and data rates up to 100~MB/s,
%\item survive and efficiently operate after $1\times10^{14} neutron eq. fluence,
%
%dosedetector components that can survive and efficiently operate in a high radiation environment with 
%local doses exceeding $XXXX$~MRad
%\end{itemize}

@@

A realistic \geant~\cite{Agostinelli2003250} simulation of the detector based on the {\sc SLIC} simulation 
package~\cite{Graf:2011zzc}
including all backgrounds has been used to optimize the layout for acceptance, tracking efficiency, mass 
resolution and prompt vertex rejection. The layout of the SVT is summarized in Tab.~\ref{tab:trk} and rendered in Fig.~\ref{fig:tracker_model}.

@@

\subsubsection{Silicon Sensors}
The sensors for the SVT are selected according to a number of requirements.  First, the sensors must be 
radiation tolerant to approximately $1.5\times10^{14}$~\fluenceunit for a six month run.  This 
corresponds to about 4~MHz/mm$^2$, so the sensor and readout technology must be capable of handling 
very high local occupancies. Since sensitivity is limited by multiple scattering, a material budget of less 
than 1\% $X_0$/layer is imperative and far less is desirable. To achieve the required vertex resolution  
single-hit resolutions better than XXXX~$\mu$m are required. Finally, the sensor technology must 
be mature and readily available at low cost.  Silicon microstrips are the best match to these requirements, and 
a supply of appropriate microstrip sensors purchased from the Hamamatsu Photonics Corporation for the 
Run 2b upgrade of the D\O experiment~\cite{Denisov:2001aa} are readily available. 
These are $p+$-on-$n$, single sided, AC coupled, 
polysilicon-biased sensors fabricated on $<100>$ silicon and have 30 (60) micron sense (readout) pitch 
over their $4\times10$~cm$^2$ surface. Although a maximum bias of 350~V is specified, the vast majority are 
operable to 1000~V and therefore remain fully depleted to a dose of approximately 
$1.5\times10^{14}$~\fluenceunit.

@@

By selecting \ee pairs from the triggered events we are able to study basic distributions of pair production 
kinematics and in particular those related to our vertex performance. Pairs of opposite charge tracks, one in the 
top and one in the bottom half of the SVT, with momentum larger than 400~MeV were selected. 
The pair production 
kinematics are relatively well reproduced given the alignment of the tracker; Fig.~\ref{fig:pair_kin} shows the 
invariant mass and ratio of electron momentum over the sum of electron and positron. 
 \begin{center}
{\small
\begin{figure*}[ht]
   \includegraphics[width=5cm]{figures/h_trk_top_px_h_trk_top_px_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
   \includegraphics[width=5cm]{figures/h_invM_h_invM_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
   \includegraphics[width=5cm]{figures/h_sumE_h_sumE_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
%   \includegraphics[scale=0.25]{test2012/vertexing/figures/h_ratioEsum_h_ratioEsum_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
\caption{Kinematic distributions for e$^+$e$^-$ pairs selected by opposite charged tracks in the top and bottom half of the tracker: track momentum in the top half of the SVT (left), invariant mass (middle) and the sum of the track momentum for the pair.}
\label{fig:pair_kin}
\end{figure*}
}
\end{center}

For the vertexing 
performance the foremost difference compared to electron beam running is that the target was 
located approximately 67~cm from our nominal target position; giving almost collinear tracks in the detector. 
This degrades the vertex resolution along the 
beam line compared to that expected in an electron beam with tracks from the nominal target position. 
Furthermore, tails of the vertex distributions are impossible to study with the finite data sample obtained 
in the test run. Nevertheless, useful information can still be 
obtained by studying the vertex distributions. Figure~\ref{fig:vtx_pos} shows the distance of closest 
approach of the momentum vectors extrapolated in the 
upstream direction from our analyzing magnet, taking into account the measured fringe field of the PS magnet. 
 \begin{center}
{\small
	 \begin{figure*}[ht]
	\includegraphics[width=5cm]{figures/h_vtx_fr_x_h_vtx_x_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\includegraphics[ width=5cm]{figures/h_vtx_fr_y_h_vtx_y_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\includegraphics[width=5cm]{figures/h_vtx_fr_z_h_vtx_z_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\caption{{\bf \textcolor{red}{NEED TO BE UPDATED}}Vertex position represented by the distance of 
	closest approach of the extrapolated momentum vectors upstream of the PS magnet taking into account 
	the measured fringe field. The overall shift from zero in the x-direction is 
	due to a 30~mrad rotation of the SVT with respect to the beam line.}
	\label{fig:vtx_pos}
\end{figure*}
}
\end{center}
While the tails of the vertex distribution expected in electron beam running is not accessible here the 
fact that the core is relatively well described provides confidence of the description of the amount of 
material and the model used for simulating multiple scattering in the target.  
These are crucial for benchmarking the HPS \Aprime physics reach since both the mass and vertex resolution 
that determine the sensitivity are limited by multiple scattering.

@@

\subsection{Trigger}
\label{sec:trigger}
The trigger system is designed to efficiently select \ee events by 
using information from the ECal . The trigger looks for time coincidences of clusters in the top and bottom half 
of the ECal.  The trigger system can be broken down into three sections (see Fig.~\ref{fig:hps_trigger_cal}):
 \begin{itemize}
 \item FADC (pulse finding): samples and digitizes the signal pulses from each detector channel. Sends the measured pulse energy and arrival time to the CTP.
\item CTP (cluster finding): groups FADC pulses from each half of the ECal into clusters. The cluster energy, arrival time, and hit pattern are sent to the SSP.
 \item SSP (cluster pair finding): searches for time coincidence of pairs of clusters from the top and bottom half of the ECal and applies topological selections.
 \end{itemize}
\begin{center}
{\small
 \begin{figure}[t]
 \includegraphics[width=7cm]{figures/hps_trigger_cal}
\caption{Block diagram of the ECAL trigger system consisting of the FADC that samples and digitizes the detector channel signals and sends them for cluster finding in the CTP. The CTP clusters are sent to the SSP where the final trigger decision is taken based on pairs of clusters in both halves of the ECal. The decision is sent to the Trigger Supervisor (TS) that generates the necessary signals to readout the sub-detectors.}
 \label{fig:hps_trigger_cal}
 \end{figure}
}
\end{center}
The time coincidence window of pairs of clusters in the top and bottom half of the ECal are programmable 
with 4~ns resolution.

@@

Occupancies close to the beam create many of the key challenges in the HPS experiment
and determine the limits of sensitivity to low \Aprime masses.
These occupancies are dominated by electrons which have multiple Coulomb scattered (MCS) 
to relatively large angles in the target. Because HPS is sensitive to scattering angles far out on the tail of the 
MCS distribution, well beyond the angles important in other experiments, care must be taken
to ensure our simulations are correct in this regime.

@@

One of the main goals of the HPS Test was to evaluate the description of the tails of the MCS
to gain further confidence in the expected detector occupancy in the full HPS experiment. 
Previous studies of multiple scattering angles show good agreement with the \moliere 
theory~\cite{Moliere:1948zz} for a wide range 
of materials and projectiles~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467}. We have verified that 
the angular distribution $F(\theta)$ in the differential cross section $d\sigma = F( \theta) d(\cos\theta) d\phi$ 
for the \egs~\cite{egs5} simulation program show good agreement with \moliere's analytical formula as 
formulated by Bethe~\cite{Bethe:1953va}. 
The small angle approximation was also shown to be in agreement with the theory formulated by Gaudsmit and 
Saunderson~\cite{Goudsmit:1940zza,Goudsmit:1940zz} that is valid for any angle. 
While \egs uses the more complex and time consuming \moliere 
formula, \geant uses an approximation with two, continuously joined, functions to take into account 
small and large angle scattering. Due to the explicit function in large angle approximation used by \geant 
we expect that \geant will overestimate the angular distribution at angles larger than a few mrad.

@@

Data was taken at three different converter thicknesses with a beam current varying between 
$30-70$~nA, see~Tab.~\ref{tab:currents}.  The photon beam line during the test run produced a relatively 
large fraction of pairs 
originating upstream of the converter. This contribution was measured during data taking 
with ``empty'' converter runs i.e. removing the converter but with all other conditions 
the same. 
The upstream background measured in the ``empty'' converter runs was subtracted 
from the other runs, properly normalized using the measured integrated currents.
\begin{table}
\begin{center}
{\small
\begin{tabular}{|c|c|c|}
\hline
Converter thickness & Duration &  $e^-$ on converter \\
 (\%r.l.) & (s) & (nC)    \\   
\hline
%\hline
1.6   & 911 &     24385.9     \\ %27 nA
%\hline
0.18   & 2640 &    193508.9  \\  %73 nA
%\hline
0.45  & 2149 &       140709.9  \\ %65.5 nA
%\hline
0    & 1279  &   88079.6  \\
\hline
\end{tabular}
\caption{Measured integrated currents for the dedicated photon runs.}
\label{tab:currents}
}
\end{center}
\end{table}

@@

This paper reviewed the HPS Test run apparatus, a simplified version of that 
planned for the full HPS experiment, and demonstrates the feasibility of the detector
technologies proposed for the silicon vertex tracker, the electromagnetic calorimeter, data acquisition and 
trigger systems and their performance. Of particular importance, data from dedicated photon beam running has 
been used to compare
the observed trigger rates with that expected in simulation. The trigger rate is almost entirely due to
photons which have converted to \ee upstream of the HPS Test setup and is sensitive to the multiple
Coulomb scattering of electrons and positrons in the conversion target.  Since scattered primary beam is 
the dominant source of occupancy in running HPS in an electron beam, good agreement between data and 
simulation confirm the background simulation used to benchmark the physics reach of the HPS experiment.

@@

In the process of developing the HPS Test design, it was found that this simple system was capable of 
delivering a surprising fraction of the physics potential anticipated for the full experiment.  With this in mind, 
we have proposed a new design for the HPS detector that builds upon the HPS Test , principally 
by addressing the compromises made for HPS Test to ensure the best possible performance for \Aprime 
physics within the envelope of the existing beam line layout and analyzing magnet. For the SVT, 
this design uses the same sensors, readout chips, and module concept and for the ECal the crystal and 
detector layout will be identical, retaining the most successful 
elements of the test run apparatus and addressing the weaknesses identified during assembly and 
operation. The new SVT layout adds a sixth layer for more robust tracking, adds acceptance in the 
three deeper layers to increase physics sensitivity and provides better silicon cooling to improve longevity. 
Other major updates needed for the full experiment are updates to the SVT electronics and 
DAQ; improving connectivity and cabling, moving digitization closer to the front-end electronics and 
updating the data reduction, event handling and trigger synchronization to reach 50~kHz trigger rates. 
This first generation of the full experiment will be ready to take physics data when CEBAF begins operation 
again in end of 2014.

>>>>

This paper reviews the HPS Test Run apparatus. It documents the successful performance of the trigger, data 
acquisition, silicon tracker, and ECal and
shows that the performance assumed in calculating the physics reach of the experiment is realistic.
Of particular importance, data from the dedicated photon beam running has been used to compare
the measured trigger rates with those expected in simulation. The Test Run trigger rate is almost entirely due to
photons which have converted to \ee{} pairs in a thin target upstream of HPS and it is sensitive to the particulars 
of their multiple Coulomb scattering in the conversion target.  Since, during electron beam running, the tails of 
the multiply Coulomb scattered primary beam are the dominant source of occupancy in the tracker and the 
trigger rate in the Ecal, they must be understood quantitatively. The trigger rates in the Test Run are also 
sensitive to the tails of the multiple Coulomb scattering distribution. Good agreement between the measured 
and simulated rates confirms our understanding of multiple Coulomb scattering and 
the reliability of the background simulation used to benchmark the physics reach of the HPS experiment.

@@

The HPS Test apparatus was designed as a 
first step of a multi-stage approach 
to the HPS experiment. The primary goal was to demonstrate that the technology and design approach 
chosen are sound for running in such a challenging environment. Nevertheless, a secondary but 
important aspect was that, after proving the technical concepts,  the apparatus would have the 
capability to discover \Aprime{} events. The implication was that the design of the HPS Test apparatus 
would be subject to almost the same constraints and requirements as the full experiment.
The kinematics of \Aprime{} production typically results in final state particles within a few degrees of the incoming beam, especially at low 
$m_{\Aprime{}}$.  
%Step
This implies that the apparatus for an \Aprime{} search in electron interactions with fix targets must operate in 
close proximity to an intense electron beam that passes through a high-Z target. 
Hence it should cope with high backgrounds and tolerate the high radiation environment. 
To meet this challenge, HPS will deploy high rate and radiation hard detector composed of 
a silicon vertex tracker with 40~MHz readout and a PbWO$_4$ crystal calorimeter with 250~MHz 
readout, capable of triggering multi-prong events at top to 50~kHz rate.

@@

The principle purpose of the HPS Test Silicon Vertex Tracker (SVT) is the efficient detection of 
charged particles and reconstruction of their momentum and angles. The precise measurements 
forms the primary input to the extraction of a signal from \Aprime{} decays; both from the precise 
measurement of the invariant mass of its \ee{} decay products and the position of its decay vertex 
downstream of the target. Track reconstruction is also important for energy scale calibration of the 
electromagnetic calorimeter due to the better momentum resolution of tracks compared to clusters 
in the calorimeter.

\subsection{SVT Requirements and Constraints}
%The primary goal was to demonstrate that the technology and design approach 
%chosen for the SVT was sound for running in the challenging environment at JLab. 
%Nevertheless, a secondary but 
%important aspect was that, after proving the technical concepts,  the apparatus would have the 
%capability to discover \Aprime{} events. The implication was that the design of the SVT in the 
%HPS Test apparatus would be subject to almost the same constraints and requirements as the full 
%experiment except for certain aspects regarding longevity and robustness subject to cost, beam time 
%and construction and design time constraints for the Test run.

@@

It is this last challenge that has the most interesting and obvious impacts on the design of the SVT. To 
achieve good vertexing performance, the first layer must be placed no more than about 10 cm downstream 
of the target. At that distance, it is found that the active region of a sensor can be placed as close to 1.5 mm 
from the center of the beam, defining a 15 mrad ``dead zone" in the detector around the plane of the 
primary beam. At the edge of this ``dead zone," the radiation dose approaches $10^{15}$ electrons/cm
$^2$/month, or roughly $3 \times 10^{14}$~\fluenceunit{}/month~\cite{Rashevskaya:2002nd}, as shown in 
Fig.~\ref{fig:radiation} and~\ref{fig:rates}.
%===================
\begin{figure}[]
\includegraphics[width=7cm]{figures/radiation}
\label{fig:radiation}
\caption{\small Simulation of the fluence 10~cm downstream of the target in the active region of the SVT ($|Y|>0.15~$cm) looking downstream; electrons bend towards $+x$.}
\end{figure}
%===================
%===================
\begin{figure}[]
\includegraphics[width=7cm]{figures/rates}
\label{fig:rates}
\caption{\small Simulation of the occupancy in the SVT 10~cm downstream of the target as a function of the 
vertical distance from the beam center.}
\end{figure}
%=================== 
This requires the sensors to be actively cooled. Meanwhile, very low-energy delta rays from beam gas interactions multiply the density of background hits, so the SVT must operate inside the beam vacuum.  Finally, in order to protect the sensors, the detector must be movable so that it can be retracted during periods of uncertain beam conditions.  
%To summarize the SVT must:
%\begin{itemize}
%\item have a uniform acceptance in the forward region; up to $\theta_y=XXXXmrad$ and as close as 
%possible to the beam axis to catch boosted decay products,
%\item >95\% reconstruction efficiency for electron and positrons  with energy down to $XXXX$~MeV,
%\item momentum resolution of XXXX\% to separate a promptly decaying \Aprime{} signal from the irreducible background,
%\item vertex resolution $<XXXX$~mm along the beam line direction to search for displaced \Aprime{} decays,
%\item event timing resolution down to 2~ns, 
%\item sustain a occupancy of up to 1\%, trigger rates of $\approx50$~kHz and data rates up to 100~MB/s,
%\item survive and efficiently operate after $1\times10^{14} neutron eq. fluence,
%
%dosedetector components that can survive and efficiently operate in a high radiation environment with 
%local doses exceeding $XXXX$~MRad
%\end{itemize}

@@

A realistic \geant{}~\cite{Agostinelli2003250} simulation of the detector based on the {\sc SLIC} simulation 
package~\cite{Graf:2011zzc}
including all backgrounds has been used to optimize the layout for acceptance, tracking efficiency, mass 
resolution and prompt vertex rejection. The layout of the SVT is summarized in Tab.~\ref{tab:trk} and rendered in Fig.~\ref{fig:tracker_model}.

@@

\subsubsection{Silicon Sensors}
The sensors for the SVT are selected according to a number of requirements.  First, the sensors must be 
radiation tolerant to approximately $1.5\times10^{14}$~\fluenceunit{} for a six month run.  This 
corresponds to about 4~MHz/mm$^2$, so the sensor and readout technology must be capable of handling 
very high local occupancies. Since sensitivity is limited by multiple scattering, a material budget of less 
than 1\% $X_0$/layer is imperative and far less is desirable. To achieve the required vertex resolution  
single-hit resolutions better than XXXX~$\mu$m are required. Finally, the sensor technology must 
be mature and readily available at low cost.  Silicon microstrips are the best match to these requirements, and 
a supply of appropriate microstrip sensors purchased from the Hamamatsu Photonics Corporation for the 
Run 2b upgrade of the D\O experiment~\cite{Denisov:2001aa} are readily available. 
These are $p+$-on-$n$, single sided, AC coupled, 
polysilicon-biased sensors fabricated on $<100>$ silicon and have 30 (60) micron sense (readout) pitch 
over their $4\times10$~cm$^2$ surface. Although a maximum bias of 350~V is specified, the vast majority are 
operable to 1000~V and therefore remain fully depleted to a dose of approximately 
$1.5\times10^{14}$~\fluenceunit{}.

@@

By selecting \ee{} pairs from the triggered events we are able to study basic distributions of pair production 
kinematics and in particular those related to our vertex performance. Pairs of opposite charge tracks, one in the 
top and one in the bottom half of the SVT, with momentum larger than 400~MeV were selected. 
The pair production 
kinematics are relatively well reproduced given the alignment of the tracker; Fig.~\ref{fig:pair_kin} shows the 
invariant mass and ratio of electron momentum over the sum of electron and positron. 
 \begin{center}
{\small
\begin{figure*}[ht]
   \includegraphics[width=5cm]{figures/h_trk_top_px_h_trk_top_px_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
   \includegraphics[width=5cm]{figures/h_invM_h_invM_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
   \includegraphics[width=5cm]{figures/h_sumE_h_sumE_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
%   \includegraphics[scale=0.25]{test2012/vertexing/figures/h_ratioEsum_h_ratioEsum_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
\caption{Kinematic distributions for e$^+$e$^-$ pairs selected by opposite charged tracks in the top and bottom half of the tracker: track momentum in the top half of the SVT (left), invariant mass (middle) and the sum of the track momentum for the pair.}
\label{fig:pair_kin}
\end{figure*}
}
\end{center}

For the vertexing 
performance the foremost difference compared to electron beam running is that the target was 
located approximately 67~cm from our nominal target position; giving almost collinear tracks in the detector. 
This degrades the vertex resolution along the 
beam line compared to that expected in an electron beam with tracks from the nominal target position. 
Furthermore, tails of the vertex distributions are impossible to study with the finite data sample obtained 
in the test run. Nevertheless, useful information can still be 
obtained by studying the vertex distributions. Figure~\ref{fig:vtx_pos} shows the distance of closest 
approach of the momentum vectors extrapolated in the 
upstream direction from our analyzing magnet, taking into account the measured fringe field of the PS magnet. 
 \begin{center}
{\small
	 \begin{figure*}[ht]
	\includegraphics[width=5cm]{figures/h_vtx_fr_x_h_vtx_x_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\includegraphics[ width=5cm]{figures/h_vtx_fr_y_h_vtx_y_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\includegraphics[width=5cm]{figures/h_vtx_fr_z_h_vtx_z_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\caption{{\bf \textcolor{red}{NEED TO BE UPDATED}}Vertex position represented by the distance of 
	closest approach of the extrapolated momentum vectors upstream of the PS magnet taking into account 
	the measured fringe field. The overall shift from zero in the x-direction is 
	due to a 30~mrad rotation of the SVT with respect to the beam line.}
	\label{fig:vtx_pos}
\end{figure*}
}
\end{center}
While the tails of the vertex distribution expected in electron beam running is not accessible here the 
fact that the core is relatively well described provides confidence of the description of the amount of 
material and the model used for simulating multiple scattering in the target.  
These are crucial for benchmarking the HPS \Aprime{} physics reach since both the mass and vertex resolution 
that determine the sensitivity are limited by multiple scattering.

@@

\subsection{Trigger}
\label{sec:trigger}
The trigger system is designed to efficiently select \ee{} events by 
using information from the ECal . The trigger looks for time coincidences of clusters in the top and bottom half 
of the ECal.  The trigger system can be broken down into three sections (see Fig.~\ref{fig:hps_trigger_cal}):
 \begin{itemize}
 \item FADC (pulse finding): samples and digitizes the signal pulses from each detector channel. Sends the measured pulse energy and arrival time to the CTP.
\item CTP (cluster finding): groups FADC pulses from each half of the ECal into clusters. The cluster energy, arrival time, and hit pattern are sent to the SSP.
 \item SSP (cluster pair finding): searches for time coincidence of pairs of clusters from the top and bottom half of the ECal and applies topological selections.
 \end{itemize}
\begin{center}
{\small
 \begin{figure}[t]
 \includegraphics[width=7cm]{figures/hps_trigger_cal}
\caption{Block diagram of the ECAL trigger system consisting of the FADC that samples and digitizes the detector channel signals and sends them for cluster finding in the CTP. The CTP clusters are sent to the SSP where the final trigger decision is taken based on pairs of clusters in both halves of the ECal. The decision is sent to the Trigger Supervisor (TS) that generates the necessary signals to readout the sub-detectors.}
 \label{fig:hps_trigger_cal}
 \end{figure}
}
\end{center}
The time coincidence window of pairs of clusters in the top and bottom half of the ECal are programmable 
with 4~ns resolution.

@@

Occupancies close to the beam create many of the key challenges in the HPS experiment
and determine the limits of sensitivity to low \Aprime{} masses.
These occupancies are dominated by electrons which have multiple Coulomb scattered (MCS) 
to relatively large angles in the target. Because HPS is sensitive to scattering angles far out on the tail of the 
MCS distribution, well beyond the angles important in other experiments, care must be taken
to ensure our simulations are correct in this regime.

@@

One of the main goals of the HPS Test was to evaluate the description of the tails of the MCS
to gain further confidence in the expected detector occupancy in the full HPS experiment. 
Previous studies of multiple scattering angles show good agreement with the \moliere{} 
theory~\cite{Moliere:1948zz} for a wide range 
of materials and projectiles~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467}. We have verified that 
the angular distribution $F(\theta)$ in the differential cross section $d\sigma = F( \theta) d(\cos\theta) d\phi$ 
for the \egs{}~\cite{egs5} simulation program show good agreement with \moliere{}'s analytical formula as 
formulated by Bethe~\cite{Bethe:1953va}. 
The small angle approximation was also shown to be in agreement with the theory formulated by Gaudsmit and 
Saunderson~\cite{Goudsmit:1940zza,Goudsmit:1940zz} that is valid for any angle. 
While \egs{} uses the more complex and time consuming \moliere{} 
formula, \geant{} uses an approximation with two, continuously joined, functions to take into account 
small and large angle scattering. Due to the explicit function in large angle approximation used by \geant{} 
we expect that \geant{} will overestimate the angular distribution at angles larger than a few mrad.

@@

Data was taken at three different converter thicknesses with a beam current varying between 
$30-70$~nA, see~Tab.~\ref{tab:currents}.  The photon beam line during the test run produced a relatively 
large fraction of pairs 
originating upstream of the converter. This contribution was measured during data taking 
with ``empty'' converter runs i.e. removing the converter but with all other conditions 
the same. 
The upstream background measured in the ``empty'' converter runs was subtracted 
from the other runs, properly normalized using the measured integrated currents.
\begin{table}
\begin{center}
{\small
\begin{tabular}{|c|c|c|}
\hline
Converter thickness & Duration &  $e^-$ on converter \\
 (\%$X_0$) & (s) & (nC)    \\   
\hline
%\hline
1.6   & 911 &     24385.9     \\ %27 nA
%\hline
0.18   & 2640 &    193508.9  \\  %73 nA
%\hline
0.45  & 2149 &       140709.9  \\ %65.5 nA
%\hline
0    & 1279  &   88079.6  \\
\hline
\end{tabular}
\caption{Measured integrated currents for the dedicated photon runs.}
\label{tab:currents}
}
\end{center}
\end{table}

@@

This paper reviewed the HPS Test run apparatus, a simplified version of that 
planned for the full HPS experiment, and demonstrates the feasibility of the detector
technologies proposed for the silicon vertex tracker, the electromagnetic calorimeter, data acquisition and 
trigger systems and their performance. Of particular importance, data from dedicated photon beam running has 
been used to compare
the observed trigger rates with that expected in simulation. The trigger rate is almost entirely due to
photons which have converted to \ee{} upstream of the HPS Test setup and is sensitive to the multiple
Coulomb scattering of electrons and positrons in the conversion target.  Since scattered primary beam is 
the dominant source of occupancy in running HPS in an electron beam, good agreement between data and 
simulation confirm the background simulation used to benchmark the physics reach of the HPS experiment.

@@

In the process of developing the HPS Test design, it was found that this simple system was capable of 
delivering a surprising fraction of the physics potential anticipated for the full experiment.  With this in mind, 
we have proposed a new design for the HPS detector that builds upon the HPS Test , principally 
by addressing the compromises made for HPS Test to ensure the best possible performance for \Aprime{} 
physics within the envelope of the existing beam line layout and analyzing magnet. For the SVT, 
this design uses the same sensors, readout chips, and module concept and for the ECal the crystal and 
detector layout will be identical, retaining the most successful 
elements of the test run apparatus and addressing the weaknesses identified during assembly and 
operation. The new SVT layout adds a sixth layer for more robust tracking, adds acceptance in the 
three deeper layers to increase physics sensitivity and provides better silicon cooling to improve longevity. 
Other major updates needed for the full experiment are updates to the SVT electronics and 
DAQ; improving connectivity and cabling, moving digitization closer to the front-end electronics and 
updating the data reduction, event handling and trigger synchronization to reach 50~kHz trigger rates. 
This first generation of the full experiment will be ready to take physics data when CEBAF begins operation 
again in end of 2014.
